---
title: "Half-life time of NICD from WB"
author:
  - name: Marek Gierlinski
    affiliations: Data Analysis Group
    note: "Creator of this report"
  - name: Hedda Meijer
    affiliations: CDB
date: today
date-format: "D MMMM YYYY"
execute:
  echo: false
  cache: true
  warning: false
format:
  html:
    theme:
      - journal
      - report.scss
    toc: true
    number-sections: true
self-contained: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: libraries
#| cache: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(forcats)
  library(targets)
  library(gt)
  library(cowplot)
})
tar_config_set(store = "../_targets")
```


```{r}
#| label: functions
N <- function(n) {
  prettyNum(n, big.mark = ",")
}

gt_table <- function(s) {
  s |> 
    gt::gt() |> 
    gt::tab_options(
      table.align = "center",
      heading.align = "center",
      data_row.padding = gt::px(3)
    ) |> 
    gt::opt_table_font(
      font = c(
        gt::default_fonts()
      )
    )
} 
```

# Proposal

We have Western Blots for Notch Intracellular Domain (NICD), with GAPDH used for loading control, at 6 time points, in 3 biological and 2-3 technical replicates, in WT and S2513A mutant. The samples were treated with LY (Gamma-secretase inhibitor) for 0-8 hours to stop production of NICD. There are three cell lines.

We want to find half-life times.

Preliminary analysis shows that after 4 hrs not much happens to WT protein levels (for WT the WB shows very little signal from 4-8 hrs).
Note, I also included data for a third cell line (Wibj2) from a separate experiment in case you have a different way to calculate half-life, please also apply this method to this dataset.
 
Questions: 
At which timepoints are the protein levels different for WT v. S2513?
Is the half-life different for WT and S2513A?
Can you please provide a short method section on the analysis? I only need a (few) sentence(s).

# Data

@fig-dat-exp shows all data, together with best-fitting exponential decay fitted to time points 0 - 4h (see below for explanation).

```{r}
#| label: fig-dat-exp
#| fig-cap: Time course of all data. The short horizontal lines show the mean of each sample. The curve represent the best-fitting exponential decay model, fitted to time points 0, 1, 2 and 4^h^ only.
#| fig-width: 8
#| fig-height: 10
tar_read(fig_dat_exp) 
```

# ANOVA and Tukey post hoc

@fig-dat-exp suggests that the Western Blot hit its sensitivity limits at about 4^h^. In S2513A and WT, the intensity flattens and in WIBJ2, there are data points missing. To assess if the exponential decay really hit a flat bottom, we perform a ANOVA followed by a post-hoc Tukey test, which is a series of pairwise t-tests corrected for multiple tests. This is done in each condition separately. The results are summarised in @fig-tukey. None of the pairs at time points 4, 6 and 8^h^ are significantly different. We conclude that the intensity after 4^h^ did not change and use only time points 0 - 4^h^ for linear regression.

```{r}
#| label: fig-tukey
#| fig-cap: Results of the Tukey post-hoc test on pairwise comparison between time points. The fill colour (diff) shows the difference between the two time points, in log~2~ scale. "x" indicates statistically significant difference at p < 0.05 level. We can see that there is no statistically significant difference between time points 4, 6 and 8^h^.
#| fig-width: 8
#| fig-height: 3
tar_read(fig_tukey)
```

# Half life

## Exponential decay model

@tbl-half-exp shows the half life time for each condition, derived from the exponential fits, together with propagated standard errors. The model used was

$$
y(t) = e^{-kt}
$$

```{r}
str_err <- function(x, std_x) {
  paste0(
      formatC(signif(x, digits = 2)),
      "±",
      formatC(signif(std_x, digits = 1))
    )
}
```

```{r}
#| label: tbl-half-exp
#| tbl-cap: Half life times from fitting exponential decay to time points 0, 1, 2, 4^h^. Errors are standard errors.
tar_read(mdl_half) |> 
   mutate(
    `k (1/h)` = str_err(k, std_k),
    `Half life (h)` = str_err(t_half, std_t_half),
  ) |> 
  select(condition, `k (1/h)`, `Half life (h)`) |> 
  gt()
```

## Exponential decay model with background

Now we take time points 4 - 8^h^ into account and include a background term in our model:

$$
y(t) = (1 - b) e^{-k t} + b
$$

Here $b$ represents a constant background, perhaps due to LY treatment not being perfect. Factor $(1 - b)$ accounts for data normalisation $y(0) = 1$.

We fit this model to the original normalised data (without log-transformation), to all time points. The best fitting curves are shown in @fig-dat-exp-b and the best-fitting parameters, including the derived half-live are shown in @tbl-half-exp-b.

```{r}
#| label: fig-dat-exp-b
#| fig-cap: Time course of all data. The short horizontal lines show the mean of each sample. The curves represent the best-fitting exponential model with background.
#| fig-width: 8
#| fig-height: 10
tar_read(fig_dat_exp_b) 
```


```{r}
#| label: tbl-half-exp-b
#| tbl-cap: Half-life times from fitting exponential model with background. Only the exponential term was used to derive the half-life. Errors are standard errors.
tar_read(mdl_b_half) |> 
  mutate(
    `k (1/h)` = str_err(k, std_k),
    b = str_err(b, std_b),
    `Half life (h)` = str_err(t_half, std_t_half),
  ) |> 
  select(condition, `k (1/h)`, b, `Half life (h)`) |> 
  gt()
```

# Comparison between conditions

Below, we compare WT with the two conditions at each time points separately, using a t-test anc correcting the results for multiple testing (Benjamini-Hochberg).

## Comparison between WT and S2513A


```{r}
#| label: fig-cmp-wt-s2513a
#| fig-cap: Results of the t-test comparison at each time point between WT and S2513A. Points and error bars represent the difference and the standard error between the two conditions, respectively. Black points and lines indicate statistically significant results (FDR < 0.05).
#| fig-width: 6
#| fig-height: 4
tar_read(fig_wt_s2513a)
```

```{r}
#| label: gt-func
gt_dif <- function(d) {
  d |> 
  gt() |> 
  fmt_number(
    columns = 2,
    n_sigfig = 2
  ) |> 
  fmt_scientific(
    columns = 3,
    rows = `Adj. p-value` < 0.01,
    n_sigfig = 2
  ) |> 
  fmt_number(
    columns = 3,
    rows = `Adj. p-value` > 0.01,
    n_sigfig = 2
  )
}
```


```{r}
#| label: tbl-cmp-wt-s2513a
#| tbl-cap: Statistical test results for comparison between WT and S2513A.
tar_read(cmp_wt_s2513a) |> 
  select(`Time (h)` = time_point, Difference = estimate, `Adj. p-value` = p_adj) |> 
  gt_dif()
```


## Comparison between WT and WIBJ2


```{r}
#| label: fig-cmp-wt-wibj2
#| fig-cap: Results of the t-test comparison at each time point between WT and WIBJ2. Points and error bars represent the difference and the standard error between the two conditions, respectively. Black points and lines indicate statistically significant results (FDR < 0.05).
#| fig-width: 6
#| fig-height: 4
tar_read(fig_wt_wibj2)
```


```{r}
#| label: tbl-cmp-wt-wibj2
#| tbl-cap: Statistical test results for comparison between WT and WIBJ2.
tar_read(cmp_wt_wibj2) |> 
  select(`Time (h)` = time_point, Difference = estimate, `Adj. p-value` = p_adj) |> 
  gt_dif()
```


# Methods

Western blot (WB) normalised intensities were log~2~ transformed. Time course plots indicated no signal change after 4^h^, likely caused by reaching detection limits of the WB. This was confirmed by ANOVA and post-hoc Tukey test for each cell line, showing no statistically significant difference (at FDR limit of 0.05) between time points 4^h^, 6^h^ and 8^h^. Therefore, only time points 0^h^-4^h^ were used for half-life calculations. To estimate half-lives, linear regression analyses were performed separately for each cell line using log~2~-transformed intensities, with regression lines constrained to pass through the origin (intercept = 0). Half-lives were calculated from the slopes of these regression lines, and standard errors were propagated accordingly. Additionally, log~2~-transformed intensities were compared between the WT and each of the two cell lines using a t-test, with p-values adjusted for multiple testing using the Benjamini-Hochberg method.


# Resources {#sec-resources}


## Code

-   [R code in GitHub](https://github.com/bartongroup/MG_NICDhalf)

# Session info {#sec-session-info}

```{r}
#| label: session-info
#| cache: false
targets::tar_load(session_info)
```

```{r}
#| label: gt_session
gt_session <- function(s) {
  s |> 
    gt::gt() |> 
    gt::tab_options(
      table.align = "left",
      data_row.padding = gt::px(3)
    )
#    gt::opt_table_font(
#      font = c(
#        gt::google_font(name = "Roboto Mono"),
#        gt::default_fonts()
#      )
#    )
}
```

### Platform

```{r}
#| label: session-info-platform
#| cache: false
session_info$platform |>
  unlist() |>
  tibble::as_tibble(rownames = "Setting") |> 
  dplyr::rename(Value = value) |> 
  gt_session() |> 
  gt::cols_width(
    Setting ~ gt::px(150)
  )
```

### Packages

```{r}
#| label: session-info-packages
#| cache: false
session_info$packages |>
  tibble::as_tibble() |>
  dplyr::filter(attached == TRUE) |> 
  dplyr::select(Package = package, Version = loadedversion, Date = date, Source = source) |> 
  gt_session() |> 
  gt::cols_width(
    Package ~ gt::px(150),
    Version ~ gt::px(80),
    Date ~ gt::px(150),
    Source ~ gt::px(400)
  )
```